---
title: "Rede_beatriz"
format: 
 html:
  toc: true
  toc-location: left
  toc-title: Sumário
  number-sections: true
  echo: true
  theme: 
   light: flatly
   dark: darkly
editor_options: 
  chunk_output_type: console
---

\![script author: Beatriz Belicha, contact: abeatrizbiomed\@gmail.com\]

# Data import

```{r}
#| message: false
#| warning: false
#| output: false

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
} 

library(tidyverse)

```

```{r}
data.cytokines <- read.csv("/Users/beatrizbelicha/Documents/bioinfo/citocina/citocinas.csv", header = T, stringsAsFactors = T)

data.cytokines <- data.cytokines |> 
  slice(-c(9, 13))
```

# Distributions and Outliers

## Define the metadata column that separates groups

Here we are assigns the string "condicao" (the name of our clinical metadata column) to the variable `condition`. This parameterizes the script. By assigning the grouping variable (e.g., controle x câncer) to an object, avoid hardcoding the column name throughout the pipeline. This makes the code dynamic and easily adaptable for future datasets.

```{r}
condition <- "condicao"
```

## Identify cytokines columns

`colnames()` extracts all column names, and `ncol()` counts the total number of columns. The `12:ncol()` indexing subsets the vector of names, selecting only those from the 12th column to the last. Early columns contain categorical metadata, we assume the actual biomarker concentrations begin at column 12. Isolating these names into `cytokines_names` allows us to apply mathematical transformations exclusively to analytical variables, preventing execution errors on demographic columns.

```{r}
cytokine_names <- colnames(data.cytokines)[12:ncol(data.cytokines)]
```

## Create empty lists to store the individual dataframes and plots

`list()` initializes empty list objects, which act as structured storage containers. To handle multiplexed data iteratively (via loops) without polluting the Global Environment. These lists act as clean folders to store individual cytokine dataframe (`df_list`) and their corresponding plots (`plot_list`) systematically during the loop execution.

```{r}
df_list <- list()
plot_list <- list()
```

## Correcting from , to .

MAGPIX software often exports numerical data with commas (e.g., 140,5) and out-of-range string (e.g., "OOR\<"). R interprets these columns as discrete characters. This specific combination rectifies the decimal formatting (140.5) and forces non-numerical strings into `NA`. This is an absolute prerequisite for rendering continuous plots and running non-parametric statistical test correctly.

The pipe `|>` forwards the raw data to `mutate`, which modifies columns. The `acroos` function applies a rule simultaneously to all columns specified in `cytokine_names`. The rule uses `gsubs(",", ".", .)` to find commas and replace them with dots, followed by `as.numeric` to coerce the result into a continuous numeric format.

```{r}
data.cytokines <- data.cytokines %>%
  mutate(across(all_of(cytokine_names), ~ as.numeric(gsub(",", ".", .))))
```

## Loop through each cytokine to subset data and generate plot

**Part 1:** The `for (cyt in cytokine_names)` statement initializes a loop. It instructs R to take the first cytokine name from the list, assign it to the placeholder `cyt`, and run the code block; then repeat for the next name, until the list ends.

- Inside the loop, a temporary dataframe named `df_temp` is created. The pipe (`%>%`) takes the main dataset (`data.cytokines`) and uses the `select`function to subset only two columns: the clinical group column (`condition`) and the current cytokine column (`cyt`).
- Finnaly, the `drop_na` function scans this two-column dataframe and removes any row (patient) containing a missing value (NA).

**Part 2:** This line takes the newly cleaned, two-column temporary dataframe (`df_temp`) and stores it inside the previously initialized list (`df_list`). The double bracket indexing `[[cyt]]` dynamically names this specific element/slot in the list using the current cytokine's string name (e.g., IL_6).

A loop is iterative. Without this lines of code, the `df_temp` object would be overwritten in the very next cycle, and the previous cytokine's cleaned data would be permanently lost. By storing each `df_temp` into `df_list`, we ensure that after the loop finishes, all subsetted and cleaned datasets are preserved and neatly organized in your RAM. If, after generating the plots, we decide to run a Shapiro-Wilk normality test exclusively on a specific cytokine, we don't need to subset the data again, we simply call `df_list[["Cytokine"]]`\` to retrieve the exact dataset.

**Part 3:** In tumor immunology, before computing any *p-value*, we must visually assess the data distribution. Cytokine-driven immune responses rely on positive feedback amplification loops. Consequently, the data rarely follows a perfect "bell curve" (Gaussian distribution). Instead, the cancer group will likely exhibit a long right-tail, reflecting biological outliers (patients with a hyper-inflamed tumor microenvironment). The transparency (`alpha`) and non-stacked overlapping (`identity`) arguments are crucial here: they allow you to see the exact translucent intersection between the basal immune tone of the control group and the pathological profile of the Gastric Cancer cohort on the same visual plane.

- `p_hist <- ggplot()`: initializes the plot canvas and assigns is to the object `p_hist`.
- `aes(...)`: The `aes` (aesthetics) function maps the visual properties. The X-axis represents the cytokine concentration. The `.data[]` syntax is an advanced tidyverse pronoun used inside loops to dynamically evaluate the string variable as a column name. The `fill` argument colors the bars according to the patient's clinical group.
- `geom_histogram(...)`: instructs ggplot to render frequency bars. `alpha = 0.6` adds transparency. `position = "identity"` forces the bars to overlap on the same baseline rather than stacking on top of each other. `bins = 30` divides the data range into 30 intervals. And `color = "black"` adds borders to the bars.
- `theme_minimal()`: strips away the default gray background, applying a clean.
- `labs(...)`: uses the `paste()` function to concatenate strings dynamically (e.g., "Distribuição da" + "IL_6"), generating the title, subtitle and axis labels.
- `scale_fill_manual(...)`: overrides default palettes with specific hexadecimal colors.

**Part 4:** In the tumor immunology, multiplexed cytokine profiles are not notoriously non-parametric. Relying on the statistical "mean" is critically flawed because it is heavily skewed by hyper-inflamed patients. The boxplot is mandatory here because it centers on the median, a robust metric resistant to extremes. Furthermore, highlighting the outliers in pink is not merely aesthetic. In clinical bioinformatics, these biological outliers are rarely technical MAGPIX errors, they often represent hyper-responsive patients. Highlighting the visually aids in generating vital biological and clinical hypotheses.

- `p_bpx <- ggplot(...)`: initializes the plot, Unlike histogram, the mapping shifts: the X-axis receives the categorical variable, and the Y-axis maps the continuous variable. `fill` colors the boxes based on the group.
- `geom_boxplot(...)`: renders the box-and-whisker plot. Unlike a standard bar chart that displays the mean, the boxplot visually represents the Median (the thick horizontal line inside the box) and the Interquartile Range (IQR) (the boundaries of the box, containing the middle 50% of our cohort).
- The outlier arguments (`outlier.color = "pink", outlier.shape = 16, outlier.size = 2`): instruct R to identify data points that fall statistically outside the whiskers and render them a solid dots (`shape = 16`), sized at `2`, and colored pink.
- `scale_fill_manual(...)`: applies our updated custom hexadecimal color palette.
- `theme(...)`: a publication-quality design tweak. Since the X-axis is already explicitly labeled with the group names, a side legend mapping colors to group is redundant. This command removes the legend, maximizing plotting area.

**Part 5:** The inherent behavior of a loop is to overwrite its internal placeholder variables. When the IL-6 iteration finishes and loop moves to IL-10, the previous `p_hist` and `p_box` objects are overwritten and lost. Without storing them in a list, once the loop finishes iterating over all our cytokines, we would only have access to the plot of the very last molecule in the array. By employing this dynamic storage strategy, we compile a comprehensive, easily accessible graphic library within our R environment. If our PI later requests a high-resolution export of a specific mediator, we won't need to rerun the pipeline, executing \`plot_list\[\["cytokine_box"\]\] will instantly retrieve and render that specific figure.

- The `paste0()` function concatenates string without inserting whitespace. It takes the variable `cyt` and merges it with the `"_hist"` or `"_box"` suffices, generating a unique identifier string, such as `"IL_6_hist"`.
- The double brackets `[[]]` use this generated string to name a specific slot within our previously initializes `plot_list`.
- The assignment operator `<-` the stores the fully rendered ggplot objects (`p_hist` and `p_box`) into these precisely named slots.

**Part 6:** The `print()` function explicitly commands R to display the `p_hist` and `p_box` graphical objects visually on our screen. If we execute the name of a ggplot object (like `p_hist`) directly in the console, it renders automatically. However, **inside a loop**, R defaults to a "silent mode" to optimize processing speed. It evaluates the plot and stores in the memory (in our list), but it suppresses visual output. Without explicitly wrapping the plot objects in the `print()` function, our script would run flawlessly, but our final document would be completely blank. This command forces the rendering engine to actively draw and embed each histogram and boxplot into our report dynamically as the loop iterates.

```{r}
for (cyt in cytokine_names) {
  #--Subsetting data--
  #Create a specific dataframe for the current cytokine, dropping missing values (NA)
  df_temp <- data.cytokines %>%
    select(all_of(condition), all_of(cyt))%>%
    drop_na()

  #Store the specific cytokine dataframe into out list
  df_list[[cyt]] <- df_temp

  #--Histogram Plot--
  #Evaluates the data distribution
  p_hist <- ggplot(df_temp, aes(x = .data[[cyt]], fill = .data[[condition]])) +
    geom_histogram(alpha = 0.6, position = "identity", bins = 30, color = "black") +
    theme_minimal()+
    labs(
      title = paste("Distribuição da", cyt),
      subtitle = "Histograma para avaliação da distribuição",
      x = "Concentração (pg/mL)",
      y = "Frequência"
    ) +
    scale_fill_manual(values = c("#e34230ff", "#345edbff")) #Blue for control, Red for cancer

  #--Boxplot--
  #Evaluates medians, interquartile ranges, and biological outliers
  p_box <- ggplot(df_temp, aes(x = .data[[condition]], y = .data[[cyt]], fill = .data[[condition]])) +
    geom_boxplot(alpha = 0.7, outlier.color = "pink", outlier.shape = 16, outlier.size = 2) +
    theme_minimal() +
    labs(
      title = paste("Nível de", cyt, "por grupo"),
      x = "Grupo Clínico",
      y = "Concentração (pg/mL)"
    ) +
    scale_fill_manual(values = c("#e34230ff", "#345edbff")) +
    theme(legend.position = "none") #Hides legend since x-axis already has the group names

  #Store plots in the list
  plot_list[[paste0(cyt, "_hist")]] <- p_hist
  plot_list[[paste0(cyt, "_box")]] <- p_box

  #Render plots in the document
  print(p_hist)
  print(p_box)

}

#In case we need to access a specific cytokine

#df_IL6 <- df_list[["IL.6"]]
```

# Testing the normality of each cytokine.

## Initialize an empty dataframe to store the final results

```{r}
normality_results <- data.frame(
  Citocina = character(),
  Condição = character(),
  p_valor = numeric(),
  Distribuição = character(),
  stringsAsFactors = FALSE
)
```

## Create the loop

Here we need to separate, because testing distinct immunological states within the same sample triggers the bimodal distribution trap, yielding false negatives. The control group and the cancer group do not belong the same biological population.

**Part 1:** The `filter()` function acts as a strict subsetting funnel. It scans the rows of our `df_temp` and creates `df_control`containing only the patients where the condition column matches "controle". It repeats this logic for the "cancer" cohort. The `filter` function guarantees biological isolation prior to testing. Furthermore, R's native `Shapiro-Wilk` function strictly requires a continuous numeric vector as its input, not a multi-variable dataframe. - The `[[cyt]]` indexing extracts the specific cytokine column from the dataframe and coerces it into a pure numeric vector (a simple array of numbers).

**Parte 2:** In the Shapiro-Wilk framework, the Null Hipothesis (H0) assumes the data is normally distributed. Statistically, a *p-value* \< 0.05 rejects this normality. Automating this logical step with `ifelse()` eliminates the need for manual interpretation of dozens of *p-values*, effectively neutralizing the risk of human error in our data pipeline. The script makes the statistical triage.

- The `shapiro.teste()` function computes the Shapiro-Wilk mathematical algorithm on our numeric vector.
- The `$p.value` suffix acts as a data extractor: the test outputs a larger list of statistics, but we command R to surgically extract only the *p-value*, storing it in `p.value_control` ou `p.value_cancer`. -The `ifelse(condition, true, false)` function is a logical decision engine. It evaluates the *p-value* threshold. If *p* \> 0.05, it outputs the string `"Normal"`. If not, it outputs `"Não Normal"`.

**Part 3:** We create a two-row temporary table (`temp_results`) containing the cytokine name, the group, the *p-value*, and the distribution status. Finally, `bind-rows`function "pastes" these two rows at the bottom of our main table (`normality_results`).

```{r}
for (cyt in cytokine_names) {
  # Retrive the cleaned dataframe for the current cytokine from our list
  df_temp <- df_list[[cyt]]

  #--Separating the groups--
  #Filter data by clinical group
  df_control <- df_temp |> filter(.data[[condition]] == "controle")
  df_cancer <- df_temp |> filter(.data[[condition]] == "cancer")

  #Extract purely the numeric vectors for the test
  vec_control <- df_control[[cyt]]
  vec_cancer <- df_cancer[[cyt]]

  #--Shapiro-Wilk Normality Test
  #Run the test and extract only p-value
  p.value_control <- shapiro.test(vec_control)$p.value
  p.value_cancer <- shapiro.test(vec_cancer)$p.value

  #Determine distribution based on the p-value threshold (Alpha = 0.05)
  dist_control <- ifelse(p.value_control > 0.05, "Normal", "Não Normal")
  dist_cancer <- ifelse(p.value_cancer > 0.05, "Normal", "Não Normal")

  #--Consolidating the results--
  temp_results <- data.frame(
    Citocina = c(cyt, cyt), 
    Condição = c("Controle", "Câncer"), 
    p_valor = c(p.value_control, p.value_cancer),
    Distribuição = c(dist_control, dist_cancer), 
    stringsAsFactors = FALSE
  )

  #Append these results to our main general dataframe
  normality_results <- bind_rows(normality_results, temp_results)
}

#Display the final table 
print(normality_results)
```

# Appling the transformation

Cytokine data commonly follows a right-skewed distribution, with most values clustered at the lower end and a few extreme high values. This violates the normality assumptions required by many parametric statistical tests.

## Log

The transformation in one the most widely used strategies to address the problem of normality, as it compresses the scale of high values and pulls the distribution closer to a normal curve. However. it's essential to verify whether the transformation was actually effective by comparing the data distribution before and after, for example, using plots or normality test. Applying a transformation without testing it can create a false sense that the assumptions have been met, ultimately compromising the validity of downstream analyses.

This code create a new datarame called `data.log` from `data.cytokines`, applying a natural logarithm transformation to all cytokine columns. The pipe operator (`|>`) passes `data.cytokines` as the first argument to the next function. The `mutate()` function, from the `dplyr` package, modifies or create columns in a dataframe, in this case, it overwrites the selected cytokines columns with their log-transformed values.

The `across()` function applies a transformation to multiple columns simultaneously, avoiding repetitive column-by-column operations. It takes two arguments:

1.  `all_of(cytokines_names)`, which selects only the columns listed in the `cytokines_names` vector and raises an error if any name is missing from the dataframe;

2.  `~log(.)`, an anonymous function where `~` defines the formula and `.` refers each column being processed, computing the natural logarithm (base *e*) of every value.

```{r}
data.log <- data.cytokines |> 
  mutate(across(all_of(cytokine_names), ~ log(.)))
```

This code creates an empty dataframe called `normality_log`, which servers as a structured container to store the results of normality tests applied to the transformed data. Each arguments defines a column with it corresponding data type:

- `Transformação = character()` and `Citocinas = character()` store, respectively, the name of the transformation applied and the name of cytokine tested. `Condição = character()` identifies the experimental group. `p_valor = numeric()` holds the *p-value* returned by the test. `Distribuição = character()` records the conclusion about the data distribution. Finally, `stringsAsFactor = FALSE` instructs R to keep text columns as character type, preventing them from being automatically converted to factor, which could cause issues when appending new rows to the dataframe.

```{r}
normality_log <- data.frame(
  Transformação = character(),
  Citocina = character(),
  Condição = character(),
  p_valor = numeric(),
  Distribuição = character(),
  stringsAsFactors = FALSE
)

```

```{r}
for (cyt in cytokine_names) {

  df_temp <- data.log %>%
    select(all_of(condition), all_of(cyt)) %>%
    drop_na()

  vec_control <- df_temp %>%
    filter(.data[[condition]] == "controle") %>%
    pull(all_of(cyt))

  vec_cancer <- df_temp %>%
    filter(.data[[condition]] == "cancer") %>%
    pull(all_of(cyt))

  p.value_control <- shapiro.test(vec_control)$p.value
  p.value_cancer <- shapiro.test(vec_cancer)$p.value

  dist_control <- ifelse(p.value_control > 0.05, "Normal", "Não Normal")
  dist_cancer <- ifelse(p.value_cancer > 0.05, "Normal", "Não Normal")

  temp_results <- data.frame(
    Transformação = "Log",
    Citocina = c(cyt, cyt),
    Condição = c("Controle", "Câncer"),
    p_valor = c(p.value_control, p.value_cancer),
    Distribuição = c(dist_control, dist_cancer),
    stringsAsFactors = FALSE
  )

  normality_log <- bind_rows(normality_log, temp_results)
}

# Display final table
print(normality_log)
```

## Log1p

The `log()` function computes the pure natural logarithm, if any value in the data is zero, the result will be `-Inf`, which breaks downstream analyses. The `log1p()` function, on the other hand, computes `log(1 + x)`, adding 1 to each value before applying the logarithm. This ensures that zeros are transformed to `log(1) = 0`, avoiding the problem entirely. For cytokines data, where zero values are biologically plausible, `log1p()` is generally the safer choice.

```{r}
data.log1p <- data.cytokines |> 
  mutate(across(all_of(cytokine_names), ~ log1p(.)))

```

```{r}
normality_log1p <- data.frame(
  Transformação = character(),
  Citocina = character(),
  Condição = character(),
  p_valor = numeric(),
  Distribuição = character(),
  stringsAsFactors = FALSE
)
```

```{r}
for (cyt in cytokine_names) {

  df_temp <- data.log1p %>%
    select(all_of(condition), all_of(cyt)) %>%
    drop_na()

  vec_control <- df_temp %>%
    filter(.data[[condition]] == "controle") %>%
    pull(all_of(cyt))

  vec_cancer <- df_temp %>%
    filter(.data[[condition]] == "cancer") %>%
    pull(all_of(cyt))

  p.value_control <- shapiro.test(vec_control)$p.value
  p.value_cancer <- shapiro.test(vec_cancer)$p.value

  dist_control <- ifelse(p.value_control > 0.05, "Normal", "Não Normal")
  dist_cancer <- ifelse(p.value_cancer > 0.05, "Normal", "Não Normal")

  temp_results <- data.frame(
    Transformação = "Log1p",
    Citocina = c(cyt, cyt),
    Condição = c("Controle", "Câncer"),
    p_valor = c(p.value_control, p.value_cancer),
    Distribuição = c(dist_control, dist_cancer),
    stringsAsFactors = FALSE
  )

  normality_log1p <- bind_rows(normality_log1p, temp_results)
}

# Display final table
print(normality_log1p)
```

```{r}
table(normality_log1p$Distribuição)
table(normality_log$Distribuição)
table(normality_results$Distribuição)
```

No significant results !!!

# PCA

## Raw data

This code creates a matrix called `pca.raw.matrix` by selecting only the cytokines columns from the dataframe `data.cytokines`. The `dplyr::select()` function is called with explicit package notation to make clear which package it comes from and to avoid conflicts with functions of the same name loaded from other packages in the session. The result is a dataframe containing only the raw, untransformed cytokines values, which will be used as input for the Principal Component Analysis (PCA).

The `prcomp()` function runs a PCA on the raw cytokines data and stores the results in the object `pca.raw`. The function is R's built-in PCA implementation and takes the following arguments:

- `pca.raw.matrix`: is the dataframe containing only the raw cytokine values, prepared in the previous step;

- `center = TRUE`: centers the data by subtracting the mean of each variable, so that all columns have a mean of zero, this is essential to prevent variables with very different means from dominating the principal components;

- `scale. = TRUE`: scales the data by dividing by the standard deviation of each variable, giving all of them a variance pf 1. This is particularly important when cytokines are measured on different scales or units, preventing those with numerically larger values from having a disproportionate influence on the analysis. The dot in the argument name is a quirk of the function designed to avoid conflict with R's base `scale()` function.

We now prepare the PCA results for visualization by organizing them into a dataframe and extracting the variance explained by each component.

`pca.raw$x` accesses the scores matrix from the `prcomp()` output, which contains the coordinates of each sample in the new dimensional space defined by the principal components. The `as.data.frame()` function converts this matrix into a dataframe, a more convenient format for plotting. A new column called `condicao` is then added to the dataframe. The `pull()` function from `dplyr` extracts a single column from a dataframe as a plain vector — in this case, the column whose name is stored in the variable `condition`. This allows each point in the PCA plot to be colored or grouped according to the experimental condition of each sample. Finally, `summary(pca.raw)$importance` returns a matrix of metrics for each principal component, where row `[2,]` corresponds to the proportion of variance explained. The `[1:2]` index selects only the first two components (PC1 and PC2), which are the ones that will be plotted. The result is multiplied by 100 to convert to a percentage and rounded to one decimal place using `round(..., 1)`.

Finally, the PCA is plotted using `ggplot2`. The `pca.raw.df` dataframe is passed as the base data, and within `aes()` the aesthetic mappings are defined: `x = PC1` and `y = PC2` position each sample on the cartesian plane according to its coordinates in the first two principal components, while `color = condicao` colors each point according to the experimental group of that sample.

- `geom_point(size = 3)` adds the points to the plot, with size 3 for better visibility.

- `labs()` defines the plot's text elements, `title` names the plot, and the `x` and `y` arguments dynamically build the axis labels using `paste0()`, which concatenates the component name with its explained variance from the `var.raw` vector, producing labels such as "PC1 (32.5%)".

- `theme_minimal()` applies a clean visual theme that removes unnecessary graphical elements, making the plot more elegant and easier to interpret.

```{r}
#Cytokines matrix

pca.raw.matrix <- data.cytokines |> 
  dplyr::select(all_of(cytokine_names))

#PCA
pca.raw <- prcomp(
  pca.raw.matrix,
  center = TRUE,
  scale. = TRUE
)

#Dataframe to plot
pca.raw.df <- as.data.frame(pca.raw$x)

#Add condition
pca.raw.df$condicao <- data.cytokines |> 
  pull(all_of(condition))

#Variance
var.raw <- round(100 * summary(pca.raw)$importance[2,1:2], 1)

#Plot
ggplot(pca.raw.df, 
aes(x = PC1,
y = PC2,
color = condicao)) +
  geom_point(size = 3) +
  labs(
    title = "PCA - Dados Brutos",
    x = paste0("PC1 (", var.raw[1], "%)"),
    y = paste0("PC2 (", var.raw[2], "%)")
  ) +
    theme_minimal()
```

## Log

```{r}
pca_log_matrix <- data.log %>%
  dplyr::select(all_of(cytokine_names))

# PCA
pca_log <- prcomp(
  pca_log_matrix,
  center = TRUE,
  scale. = TRUE
)

# dataframe
pca_log_df <- as.data.frame(pca_log$x)

# condição
pca_log_df$condicao <- data.log %>%
  pull(all_of(condition))

# variância explicada
var_log <- round(100 * summary(pca_log)$importance[2,1:2], 1)

# gráfico
ggplot(pca_log_df,
       aes(x = PC1,
           y = PC2,
           color = condicao)) +

  geom_point(size = 3) +

  labs(
    title = "PCA - Dados Log",
    x = paste0("PC1 (", var_log[1], "%)"),
    y = paste0("PC2 (", var_log[2], "%)")
  ) +

  theme_minimal()
```

## Log1p

```{r}

pca_log1p_matrix <- data.log1p %>%
  dplyr::select(all_of(cytokine_names))

# PCA
pca_log1p <- prcomp(
  pca_log1p_matrix,
  center = TRUE,
  scale. = TRUE
)

# dataframe
pca_log1p_df <- as.data.frame(pca_log1p$x)

# condição
pca_log1p_df$condicao <- data.log1p %>%
  pull(all_of(condition))

# variância explicada
var_log1p <- round(100 * summary(pca_log1p)$importance[2,1:2], 1)

# gráfico
ggplot(pca_log1p_df,
       aes(x = PC1,
           y = PC2,
           color = condicao)) +

  geom_point(size = 3) +

  labs(
    title = "PCA - Dados Log1p",
    x = paste0("PC1 (", var_log1p[1], "%)"),
    y = paste0("PC2 (", var_log1p[2], "%)")
  ) +

  theme_minimal()
```

# Mann-Whitney

```{r}
#| message: false
#| warning: false
#| output: false

if (!requireNamespace("rstatix", quietly = TRUE)) {
  install.packages("rstatix")
}

library(rstatix)
```

## Raw data

```{r}
results.mann.raw <- data.frame(
  Citocina = character(),
  p_value = numeric(),
  Resultado = character()
)
```

This block runs the Mann-Whitney test for each cytokine individually and stores the results in a dataframe. The `for(cyt in cytokine_names)` loop iterates over each cytokine name in the `cytokine_names` vector, with `cyt` taking a different cytokine name at each iteration. Inside the loop, `as.formula(paste(cyt, "~", condition))` dynamically builds the statistical formula by concatenating the cytokine name, the tilde `~`, and the condition name into a string such as `"IL6 ~ group"`, then converting it into a formula object that R can interpret statistically.

`wilcox.test()` runs the Mann-Whitney test (also known as the Wilcox rank-sum test) comparing cytokine values between the groups defined by condition variable. It is a non-parametric test, appropriate for data that does not follow normal distribution.

At each iteration, a temporary dataframe `temp` is created with three columns:

1.  The cytokine name;

2.  The *p-value*;

3.  Classification using `ifelse()`.

The `ifelse()` function returns "Significativo" if the *p-value* is less than or equal to 0.05 and "Não Significativo" otherwise. The `bind_rows()` function from `dplyr` the appends temp to the main results `results.mann.raw` dataframe, accumulating results across iterations.

After the loop, `print()` displays the complete results dataframe, and `table()` produces a quick count of how many cytokines were classified as significant or non-significant.

```{r}
for(cyt in cytokine_names){
  #--Creating the standard formula--
  formula.test <- as.formula(
    paste(cyt, "~", condition)
  )
  #--Running Mann-Whitney--
  test <- wilcox.test(
    formula.test, data.cytokines
  )
  #--Saving the results--
  temp <- data.frame(
    Citocina = cyt,
    p_value = test$p.value, 
    Resultado = ifelse(test$p.value > 0.05, "Não Significativo", "Significativo")
  )
  #Append these results to our main general dataframe
  results.mann.raw <- bind_rows(
    results.mann.raw,
    temp
  )
}

print(results.mann.raw)
table(results.mann.raw$Resultado)
```

## Log

```{r}
results.mann.log <- data.frame(
  Citocina = character(),
  p_value = numeric(),
  Resultado = character()
)
```

```{r}
for(cyt in cytokine_names){
  #--Creating the standard formula--
  formula.test <- as.formula(
    paste(cyt, "~", condition)
  )
  #--Running Mann-Whitney--
  test <- wilcox.test(
    formula.test, data.log
  )
  #--Saving the results--
  temp <- data.frame(
    Citocina = cyt,
    p_value = test$p.value, 
    Resultado = ifelse(test$p.value > 0.05, "Não Significativo", "Significativo")
  )
  #Append these results to our main general dataframe
  results.mann.log <- bind_rows(
    results.mann.log,
    temp
  )
}

print(results.mann.log)
table(results.mann.log$Resultado)
```

This block applies a correction for multiple comparisons to the Mann-Whitney test results and updates the significance classification accordingly.

`p.adjust(results.mann.log$p_value, method = "fdr")` adjusts the raw p-values using the Benjamini-Hochberg method, also known as the False Discovery Rate (FDR) correction. When multiple statistical tests are performed simultaneously — one for each cytokine — the probability of obtaining at least one false positive increases with each additional test. The FDR method controls the expected proportion of false positives among the results classified as significant, producing adjusted p-values that account for the total number of tests performed. These adjusted values are stored in a new column called `FDR` in the `results.mann.log` dataframe.

`ifelse(results.mann.log$FDR < 0.05, "Sim", "Não")` then reclassifies each cytokine as significant or not based on the adjusted p-value rather than the raw one, which is the correct approach after applying a multiple comparisons correction. The result is stored in a new column called `Significativo_FDR`. It is common to find that some cytokines that appeared significant based on raw p-values lose their significance after FDR correction — this is expected and is precisely the purpose of the adjustment.

```{r}
results.mann.log$FDR <- p.adjust(
  results.mann.log$p_value,
  method = "fdr"
)

results.mann.log$Significativo_FDR <- ifelse(
  results.mann.log$FDR < 0.05,
  "Sim",
  "Não"
)
```

IL.2, IL.6, IL.8, IL.10 and GCSF remained significant.

# Welch's t-test for normal cytokines

```{r}
normal.cytokines <- c("IL.17", "IL.1B", "IL.2", "IL.13", "IL.15", "Eotaxin", "GCSF", "MIP1a", "MIP1b")
```

```{r}
results.ttest <- data.frame(
  Citocina = character(),
  p_value = numeric(),
  Resultado = character(),
  stringsAsFactors = FALSE
)
```

```{r}
for(cyt in normal.cytokines){

  formula.test <- as.formula(
    paste(cyt, "~", condition)
  )

  test <- t.test(
    formula.test, data.log,
  )

  temp <- data.frame(
    Citocina = cyt,
    p_value = test$p.value,
    Resultado = ifelse(
      test$p.value < 0.05,
      "Significativo",
      "Não significativo"
    )
  )

  results.ttest <- bind_rows(
    results.ttest,
    temp
  )
}

print(results.ttest)
table(results.ttest$Resultado)
```

As a result, IL.15, which was previously classified as "Não Significativo", has now become "Significativo". The remaining results remained unchanged.

# Heatmap

```{r}
#| message: false
#| warning: false
#| output: false

if (!requireNamespace("pheatmap", quietly = TRUE)) {
  install.packages("pheatmap")
}

library(pheatmap)
```

This code prepares the data matrix that will be used to generate the heatmap. The pipe chains three operations on `data.log`:

1.  `dplyr::select(all_of(cytokine_names))` selects only the cytokines columns;

2.  `as.matrix()` then converts the resulting dataframe into a matrix. This conversion is necessary because `pheatmap()` function require the data to be in matrix format rather than a dataframe;

```{r}
heatmap.matrix <- data.log |> 
  dplyr::select(all_of(cytokine_names)) |> 
  as.matrix()
```

This code assigns names to the rows of the `heatmap.matrix`. The `rownames()` function defines or replaces the row names of matrix or dataframe. On the right-hand side, `paste0()` concatenates the string `"amostra_"` with a numeric sequence from 1 to the total number of rows in the matrix, generate by `1:nrow(heatmap.matrix)`. This produces labels such as `"amostra_1"`, and so on. These labels are important for heatmap visualization, as they identify each sample on the corresponding axis of the plot.

```{r}
rownames(heatmap.matrix) <- paste0("amostra_", 1:nrow(heatmap.matrix))
```

This code creates an annotation dataframe that will be used to color the samples in the heatmap according to their experimental group.

`data.frame(Grupo = data.log[[condition]]` creates a dataframe with a single column called `Grupo`. The `[[condition]]` operator extracts the column whose name is stored in the variable `condition` from `data.log` as a plain vector, which is then assigned to the Grupo column. It's worth noting the difference between `[[]]` and `$`: the `$` operator requires the column name to be written directly in the code, whereas `[[]]` accepts a variable containing the name, making the code more flexible and reusable.

`rownames(annotation.df) <– rownames(heatmap.matrix)` assigns the same row names from the heatmap matrix to the annotation dataframe. This is essential for the heatmap function to correctly align each sample with its corresponding group annotation.

```{r}
annotation.df <- data.frame(
  Grupo = data.log[[condition]]
)

rownames(annotation.df) <- rownames(heatmap.matrix)
```

This code creates a list called `annotation_colors` that defines the colors used to represent each group in the heatmap. In R, lists are structures that can hold elements of different types and sizes, and this is the format expected by functions such as `pheatmap()` for receiving annotation groups.

The `Grupo` element of the list is a named color vector in hexadecimal format, where each name corresponds to a level of the experimental group: `cancer` will be displayed in red `"#e34230ff"` and `controle` in blue `"#345edbff"`. The name `Grupo` must be identical to the column name in `annotation.df`, as this is how the heatmap function matches the annotation to its corresponding colors.

```{r}
annotation_colors <- list(
  Grupo = c(
    cancer = "#e34230ff",
    controle = "#345edbff"
  )
)
```

This code generates the heatmap using pheatmap() function from the package of the same name. each argument controls a specific aspect of the visualization:

- t(heatmap.matrix) transpose the matrix, swapping rows and columns so that cytokines appear as rows and samples as columns, which is the most common and intuitive orientation for biological heatmaps. `scale = "row"` normalizes the data by row, transforming the values of each cytokine into z-scores by subtracting the mean and dividing by the standard deviation. This allows the expression pattern of cytokines with very different magnitudes to be compared without one visually dominating the others.

- `annotation_col = annotation.df` adds an annotation bar at the top of the columns, coloring each sample according to its experimental group using the annotation dataframe created earlier. `annotation_colors = annotation_colors` assigns the specific colors for each group in that annotation bar, using the color list defined previously.

- `clustering_method = "complete"` sets the hierarchical clustering method used to order both samples and cytokines in the heatmap. The `"complete"` method calculates the distance between two clusters as the greatest distance between any two points belonging to them, tending to produce more compact and well-defined clusters. `border_color = "#fcfcfc"` sets the border color between heatmap cells to a near-white tone, creating a subtle separation without distracting from the visualization. Finally, `fontsize = 10` sets the font size of the labels in the plot.

```{r}
pheatmap(
  t(heatmap.matrix),
  scale = "row",
  annotation_col = annotation.df,
  annotation_colors = annotation_colors,
  clustering_method = "complete", 
  border_color = "#fcfcfc",
  color = colorRampPalette( c("#345edbff", "white", "#e34230ff"))(100),
  fontsize = 10
)
```

## Heatmap with significant cytokines

```{r}
significant_cytokines <- c(
  "RANTES",
  "IL.2",
  "IL.6",
  "IL.8",
  "IL.10",
  "IL.11",
  "FGFbasic",
  "GCSF",
  "MPC.1", 
  "IL.13"
)
```

```{r}
heatmap.sig <- data.log |>
  dplyr::select(all_of(significant_cytokines)) |>
  as.matrix()
```

```{r}
rownames(heatmap.sig) <- paste0(
  "amostra_",
  1:nrow(heatmap.sig)
)
```

```{r}
annotation.df <- data.frame(
  Grupo = data.log[[condition]]
)

rownames(annotation.df) <- rownames(heatmap.sig)
```

```{r}
pheatmap(
  t(heatmap.sig),
  scale = "row",
  annotation_col = annotation.df,
  annotation_colors = annotation_colors,
  clustering_method = "complete", 
  border_color = "#fcfcfc",
  color = colorRampPalette(
    c("#345edbff", "white", "#e34230ff")
  )(100),
  fontsize = 10
)
```

# Cytokine-cytokine correlation

He we creates a new dataframe called `df.cancer` containing only the samples from the cancer group, keeping solely the cytokines columns. The pipe chains three operations on `data.cytokine`:

1.  `filter(condicao == "cancer")` filters the rows of the dataframe, retaining only those where the `condicao` column is equal to `"cancer"`, using the `==` operator for exact string matching;
2.  `select(all_off(cytokine_names))` the select only the cytokine columns, removing the condition column and any other variable not relevant to the correlation analysis.

```{r}
df.cancer <- data.cytokines |> 
  filter(condicao == "cancer") |> 
  select(all_of(cytokine_names))
```

This code computes the correlation matrix between cytokines using the Spearman correlation coefficient. The `cor()` function takes `df.cancer` and returns a square matrix where each cell represents the correlation between a pair of cytokines. The `method = "spearman"` argument specifies the use of Spearman correlation rather than Pearson. \
\
Spearman correlation coefficient is the most appropriate choice in this case, as it is based on the ranked values of the data and does not assume normality, making it more robust for biological data that frequently presents asymmetric distributions or outliers.

```{r}
cancer.matrix <- cor(
  df.cancer, 
  method = "spearman"
)

pheatmap(cancer.matrix)

df_cancer_matrix <- as.data.frame(cancer.matrix)
```

This loop iterates over each cytokine in the correlation matrix and creates an individual dataframe containing only the cytokines that show a strong correlation (above 0.8) with it.

1.  `for(cyt in colnames(cancer.matrix))` iterates over the column names of the correlation matrix, where each column represents a cytokine. `cor.value <- cancer.matrix[, cyt]` extracts the column corresponding to the current cytokine, retrieving its correlation coefficients with all other cytokines.
2.  `data.frame(Citocina = names(cor.value), Coeficiente = round(as.numeric(cor.value), 2))` creates a temporary dataframe with two columns: the name of each cytokine and its correlation coefficient with the current cytokine, rounded to two decimal places. `as.numeric()` ensures the values are treated as numbers.
3.  `filter(Coeficiente > 0.8 & Citocina != cyt)` retains only cytokines with a correlation above 0.8, excluding the cytokine being compared against itself — since the correlation of a variable with itself is always 1 and carries no useful information.

Finally, `assign(paste0("df.", cyt), df.filter)` dynamically creates an object in the R environment with a name built by concatenating `"df."` with the cytokine name — for example, `df.IL6` or `df.TNF`. This allows each cytokine to have its own dataframe of strong correlations without requiring manual naming.

```{r}
for (cyt in colnames(cancer.matrix)) {
  cor.value <- cancer.matrix[, cyt]
  df_temp <- data.frame(
    Citocina = names(cor.value),
    Coeficiente = round(as.numeric(cor.value), 2),
    stringsAsFactors = FALSE
  )
  df.filter <- df_temp |> 
    filter(Coeficiente > 0.8 & Citocina != cyt)
  df.name <- paste0("df.", cyt)
  assign(df.name, df.filter)
}
```

# ROC curve

```{r}
#| message: false
#| warning: false
#| output: false

if (!requireNamespace("pROC", quietly = TRUE)) {
  install.packages("pROC")
}

library(pROC)
```

```{r}
auc.table <- data.frame(
  Citocina = character(),
  AUC = numeric(),
  stringsAsFactors = FALSE
)

roc.curve <- list()
```

```{r}
for (cyt in significant_cytokines) {
  #--Calculate the ROC object mathematically--
  roc_curve <- roc(response = data.cytokines[[condition]],
  predictor = data.cytokines[[cyt]], 
  quiet = TRUE) #prevents R from flooding the console with messages

  #--Extract AUC rounded to 2 decimal places (just like your reference image)--
  auc_val <- round(auc(roc_curve), 2)

  #--Plot using the native pROC plot function to match the requested aesthetic--
  plot(roc_curve,
       main = cyt,                                    # Title (Cytokine name)
       col = "blue",                                  # Blue curve
       lwd = 2,                                       # Line thickness
       xlab = paste0("Specificity; auc=", auc_val),   # Custom X-axis label with AUC
       ylab = "Sensitivity",                          # Y-axis label
       legacy.axes = FALSE,                           # Keeps X-axis from 1.0 to 0.0
       identity.col = "darkgrey",                     # Grey diagonal line
       identity.lty = 1,                              # Solid diagonal line (1) instead of dashed (2)
       print.auc = FALSE,                             # Prevents printing AUC inside the plot area
       xlim = c(1, 0), 
       xaxs = "i", 
       yaxs = "i")
}

par(mfrow = c(1, 1), pty = "m")
```

```{r}
for (cyt in significant_cytokines) {
  #--Calculate the ROC object mathematically--
  roc_curve <- roc(response = data.cytokines[[condition]],
  predictor = data.cytokines[[cyt]], 
  quiet = TRUE) #prevents R from flooding the console with messages
  
  #--Store the curve format in our list later plotting--
  roc.curve[[cyt]] <- roc_curve

  #--Extract the mathematical AUC value rounded to 3 decimal places--
  temp.auc <- data.frame(
    Citocina = cyt, 
    AUC = round(auc(roc_curve), 2), 
    stringsAsFactors = FALSE
  )

  #--Append the result to our final table--
  auc.table <- bind_rows(auc.table, temp.auc)
}
```


## Ranking and visualization 

```{r}
auc.table <- auc.table |> 
  arrange(desc(AUC))

print(auc.table)
```

```{r}
p.roc.comb <- ggroc(roc.curve, alpha = 0.8, size = 1) +
    theme_minimal() +
    labs(
      title = "Performance Diagnóstica das Citocinas (ROC)",
      x = "Especificidade", 
      y = "Sensibilidade", 
      color = "Biomarcadores"
    ) +
      #Add the 50% diagonal line (Absolute chance)
    geom_abline(intercept = 1, slope = 1, linetype = "dashed", color = "gray50", size = 0.8)

print(p.roc.comb)
```

# TNM analysis

```{r}
data.cancer <- data.cytokines |> 
  filter(condicao == "cancer")
```

```{r}
mann.TNM <- data.frame(
  Citocina = character(),
  p_value = numeric(),
  Resultado = character(),
  stringsAsFactors = FALSE
)
TNM <- "TNM"

```
```{r}
for(cyt in cytokine_names){ 

  #--Creating the standard formula-- 
  formula.test <- as.formula( paste(cyt, "~", TNM) ) 

  #--Running Mann-Whitney-- 
  test <- wilcox.test( formula.test, data.cancer, exact = FALSE ) 

  #--Saving the results-- 
  temp <- data.frame( 
    Citocina = cyt, 
    p_value = test$p.value, 
    Resultado = ifelse(test$p.value > 0.05, "Não Significativo", "Significativo") ) 

  #Append these results to our main general dataframe 
  mann.lauren <- bind_rows( mann.TNM, temp ) }

```

# Lauren's Classification 

```{r}
mann.lauren <- data.frame(
  Citocina = character(),
  p_value = numeric(),
  Resultado = character(),
  stringsAsFactors = FALSE
)
subclass <- "subclass"

```
```{r}
for(cyt in cytokine_names){ 

  #--Creating the standard formula-- 
  formula.test <- as.formula( paste(cyt, "~", subclass) ) 

  #--Running Mann-Whitney-- 
  test <- wilcox.test( formula.test, data.cancer, exact = FALSE ) 

  #--Saving the results-- 
  temp <- data.frame( 
    Citocina = cyt, 
    p_value = test$p.value, 
    Resultado = ifelse(test$p.value > 0.05, "Não Significativo", "Significativo") ) 

  #Append these results to our main general dataframe 
  mann.lauren <- bind_rows( mann.lauren, temp ) }

```
